Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-06-22 forecast/paper/paper-medrxiv.tex
Add medRXiv submission doc; fix Addison's middle initial

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

@@

% Please give the surname of the lead author for the running footer
\leadauthor{McDonald}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Forecasting is a critical element in the public health
  response to any fast-moving epidemic or pandemic.  Predicting the future
  spread of some disease process, based on disease activity in the recent
  past---say, using what is known as autoregressive modeling in the time series
  literature---is a basic and widely-used method.  In this paper we study
  whether including auxiliary indicators as additional features in such an
  autoregressive model can be useful in COVID-19 case forecasting and hotspot
  prediction.  We find that indicators based on de-identified medical insurance 
  claims, self-reported symptoms through online surveys, and COVID-related
  internet search activity all provide a nontrivial boost in accuracy, though
  their utility generally depends the pandemic's local dynamics (upswing,
  downswing, or flat period) at prediction time.
}

% Please include corresponding author, author contribution and author
% declaration information

\authorcontributions{D.J.M., J.B., A.G., A.J.H., and R.J.T. wrote the manuscript,
produced figures, and developed software necessary to perform the analyses. All
authors performed research and provided intellectual contributions. R.J.T.
acquired funding.}
\authordeclaration{The authors have no competing interests to declare.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be
  addressed. E-mail: daniel@stat.ubc.ca} 
\equalauthors{\textsuperscript{2}J.B., A.G., and A.J.H. contributed equally to 
  this work.}

% At least three keywords are required at submission. Please provide three to
% five keywords, separated by the pipe symbol. 
\keywords{COVID-19 $|$ forecasting $|$ hotspot prediction $|$ time series | 
  digital surveillance}

\begin{abstract}
  Reliable, short-term forecasts of traditional public health reporting streams
  (such as cases, hospitalizations, and deaths) are a key ingredient in
  effective public health decision-making during a pandemic. Since April 2020,
  our research group has worked with data partners to collect, curate, and make
  publicly available numerous real-time COVID-19 indicators, providing multiple
  views of pandemic activity.  This paper studies the utility of these
  indicators from a forecasting perspective. We focus on five indicators,
  derived from medical insurance claims data, web search queries, and online
  survey responses. For each indicator, we ask whether its inclusion in a simple
  model leads to improved predictive accuracy relative to a similar model
  excluding it. We consider both probabilistic forecasting of confirmed COVID-19 
  case rates and binary prediction of case ``hotspots''.  Since the values of 
  indicators (and case rates) are commonly revised over time, we take special
  care to ensure that the data provided to a forecaster is the version that
  would have been available at the time the forecast was made.   
  % We demonstrate how less careful backtesting can lead to misleading
  % retrospective evaluations.   
  Our  analysis shows that consistent but modest gains in predictive accuracy 
  are obtained by using these indicators, and furthermore, these gains are
  related to periods in which the auxiliary indicators behave as ``leading
  indicators'' of case rates.     
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\dropcap{T}racking and forecasting indicators from public health reporting
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the utility provided by a core set of these indicators for two
fundamental prediction tasks: probabilistic forecasting of COVID-19 case 
rates and prediction of future COVID-19 case hotspots (defined by the event that
a relative increase in COVID-19 cases exceeds a certain threshold).

@@

\begin{figure}[tb!]
  \includegraphics[width=\columnwidth]{fig/revisions.pdf}
  % RJT: Ensuring we start with a "declarative title" :) 
  % https://www.internationalscienceediting.com/how-to-write-a-figure-caption/
  \caption{Revision behavior for two indicators in the HRR containing Charlotte, 
    North Carolina.  Each colored line corresponds to the data as reported on a
    particular date (\textit{as of} dates varying from September 28 through
    October 19). The left panel shows the DV-CLI signal, which was regularly
    revised throughout the period, though the effects fade as we look further
    back in time. In contrast, the right panel shows case rates reported by JHU
    CSSE (smoothed with a 7-day trailing average), which remain ``as reported''
    on September 28, with a spike towards the end of this period, until a major 
    correction is made on October 19, which brings this down and affects all
    prior data as well.}  
  \label{fig:vintage}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption{Summary of forecasting and hotspot prediction tasks considered in
  this paper.}
\begin{tabular}{l p{2.75in} p{2.75in}}
  \toprule
  & \textbf{Forecasting} & \textbf{Hotspot prediction} \\
  \midrule
  Response variable & $Y_{\ell,t}$ (7-day trailing average of COVID-19 case 
incidence rates, per location $\ell$ and time $t$) & $Z_{\ell,t} =
\indicator{Y_{\ell,t} \geq 1.25 \cdot Y_{\ell,t-7}}$ (indicator that 
$Y_{\ell,t}$ grows by more than 25\% relative to the preceding week) \\ 
  Geographic resolution & Hospital referral region (HRR) & Hospital referral
region (HRR) \\ 
  Forecast period & June 9--December 31, 2020 & June 16--December 31, 2020 \\  
  Model type & Quantile regression & Logistic regression \\
  Evaluation metric & Weighted interval score (WIS) & Area under curve (AUC) \\
  \bottomrule
\end{tabular}
\label{tab:analysis_tasks}
\end{table*}

@@

In addition to this pure autoregressive model, we also consider five
probabilistic forecasting models of the form:  
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \ | \ Y_{\ell,s},\ X_{\ell,s},\ s \leq t) 
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary indicators as we do for the case
rates. Training then proceeds just as before: we use the same 7 quantile levels
in \eqref{eq:quantile_levels}, and fit quantile regression separately for each 
level $\tau$, using data from all 306 HRRs and a trailing window of 21 days of
training data.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{fig/trajectory.pdf}
\caption{Forecast for the HRR containing New York City from an autoregressive
  model made on October 15 (vertical line).  The fan displays 50\%, 80\% and
  95\% intervals while the orange curve shows the median forecast. The black
  curve shows ``finalized'' data, as reported in May 2021.}   
\label{fig:trajectory}
\end{figure}

@@

In addition to this pure autoregressive model, we also consider five
logistic models of the form:  
\begin{multline}
\label{eq:hotspot_ar_x}
\logit \big( \P(Z_{\ell,t+a} = 1 \ | \ Y_{\ell,s},\ X_{\ell,s},\ s \leq t) \big)
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit the above models,
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x}, using logistic regression,   
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/leading-only-1.pdf}
  \caption{Correlation of the difference in WIS with the ``leadingness'' of the
    indicator at the target date, stratified by up, down, or flat period.} 
  \label{fig:leading}
\end{figure}

@@

\acknow{We thank Matthew Biggerstaff, Logan Brooks, Johannes Bracher, Michael 
  Johansson, Evan Ray, Nicholas Reich, and Roni Rosenfeld for several
  enlightening conversations about forecasting, scoring, and evaluation.  
  This material is based on work supported by gifts from Facebook, Google.org,
  the McCune Foundation, and Optum; Centers for Disease Control and Prevention
  (CDC) grant U01IP001121; the Canadian Statistical Sciences Institute; National 
  Sciences and Engineering Research Council of Canada (NSERC) grant
  RGPIN-2021-02618; and National Science Foundation Graduate Research
  Fellowship Program (NSF GRFP) award DGE1745016.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{../../common/covidcast.bib}

>>>>

\documentclass{article}
% Use the lineno option to display guide line numbers if required.

% \templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\usepackage{authblk}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage[numbers,square,sort&compress]{natbib}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}

@@

% Please add a significance statement to explain the relevance of your work
% \significancestatement{Forecasting is a critical element in the public health
%   response to any fast-moving epidemic or pandemic.  Predicting the future
%   spread of some disease process, based on disease activity in the recent
%   past---say, using what is known as autoregressive modeling in the time series
%   literature---is a basic and widely-used method.  In this paper we study
%   whether including auxiliary indicators as additional features in such an
%   autoregressive model can be useful in COVID-19 case forecasting and hotspot
%   prediction.  We find that indicators based on de-identified medical insurance 
%   claims, self-reported symptoms through online surveys, and COVID-related
%   internet search activity all provide a nontrivial boost in accuracy, though
%   their utility generally depends the pandemic's local dynamics (upswing,
%   downswing, or flat period) at prediction time.
% }

% At least three keywords are required at submission. Please provide three to
% five keywords, separated by the pipe symbol.

@@

\maketitle

\begin{abstract}
  Reliable, short-term forecasts of traditional public health reporting streams
  (such as cases, hospitalizations, and deaths) are a key ingredient in
  effective public health decision-making during a pandemic. Since April 2020,
  our research group has worked with data partners to collect, curate, and make
  publicly available numerous real-time COVID-19 indicators, providing multiple
  views of pandemic activity.  This paper studies the utility of these
  indicators from a forecasting perspective. We focus on five indicators,
  derived from medical insurance claims data, web search queries, and online
  survey responses. For each indicator, we ask whether its inclusion in a simple
  model leads to improved predictive accuracy relative to a similar model
  excluding it. We consider both probabilistic forecasting of confirmed COVID-19
  case rates and binary prediction of case ``hotspots''.  Since the values of 
  indicators (and case rates) are commonly revised over time, we take special
  care to ensure that the data provided to a forecaster is the version that
  would have been available at the time the forecast was made.   
  % We demonstrate how less careful backtesting can lead to misleading
  % retrospective evaluations.   
  Our  analysis shows that consistent but modest gains in predictive accuracy 
  are obtained by using these indicators, and furthermore, these gains are
  related to periods in which the auxiliary indicators behave as ``leading
  indicators'' of case rates.

  {\bf Keywords:} COVID-19 $|$ forecasting $|$ hotspot prediction $|$ time series
    $|$ digital surveillance
\end{abstract}

\footnotetext[1]{To whom correspondence should be
  addressed. E-mail: \href{mailto:daniel@stat.ubc.ca}{daniel@stat.ubc.ca}} 
\footnotetext[2]{J.B., A.G., and A.J.H. contributed equally to 
  this work.}

\clearpage

Tracking and forecasting indicators from public health reporting
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the utility provided by a core set of these indicators for two
fundamental prediction tasks: probabilistic forecasting of COVID-19 case 
rates and prediction of future COVID-19 case hotspots (defined by the event that
a relative increase in COVID-19 cases exceeds a certain threshold).

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=.5\linewidth]{fig/revisions.pdf}
  % RJT: Ensuring we start with a "declarative title" :) 
  % https://www.internationalscienceediting.com/how-to-write-a-figure-caption/
  \caption{Revision behavior for two indicators in the HRR containing Charlotte, 
    North Carolina.  Each colored line corresponds to the data as reported on a
    particular date (\textit{as of} dates varying from September 28 through
    October 19). The left panel shows the DV-CLI signal, which was regularly
    revised throughout the period, though the effects fade as we look further
    back in time. In contrast, the right panel shows case rates reported by JHU
    CSSE (smoothed with a 7-day trailing average), which remain ``as reported''
    on September 28, with a spike towards the end of this period, until a major 
    correction is made on October 19, which brings this down and affects all
    prior data as well.}  
  \label{fig:vintage}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption{Summary of forecasting and hotspot prediction tasks considered in
  this paper.}
\begin{tabular}{l p{2.25in} p{2.25in}}
  \toprule
  & \textbf{Forecasting} & \textbf{Hotspot prediction} \\
  \midrule
  Response variable & $Y_{\ell,t}$ (7-day trailing average of COVID-19 case 
incidence rates, per location $\ell$ and time $t$) & $Z_{\ell,t} =
\indicator{Y_{\ell,t} \geq 1.25 \cdot Y_{\ell,t-7}}$ (indicator that 
$Y_{\ell,t}$ grows by more than 25\% relative to the preceding week) \\ 
  Geographic resolution & Hospital referral region (HRR) & Hospital referral
region (HRR) \\ 
  Forecast period & June 9--December 31, 2020 & June 16--December 31, 2020 \\  
  Model type & Quantile regression & Logistic regression \\
  Evaluation metric & Weighted interval score (WIS) & Area under curve (AUC) \\
  \bottomrule
\end{tabular}
\label{tab:analysis_tasks}
\end{table*}

@@

In addition to this pure autoregressive model, we also consider five
probabilistic forecasting models of the form:  
\begin{equation}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \ | \ Y_{\ell,s},\ X_{\ell,s},\ s \leq t) 
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{equation}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary indicators as we do for the case
rates. Training then proceeds just as before: we use the same 7 quantile levels
in \eqref{eq:quantile_levels}, and fit quantile regression separately for each 
level $\tau$, using data from all 306 HRRs and a trailing window of 21 days of
training data.

@@

\begin{figure}[t]
  \centering
\includegraphics[width=.5\columnwidth]{fig/trajectory.pdf}
\caption{Forecast for the HRR containing New York City from an autoregressive
  model made on October 15 (vertical line).  The fan displays 50\%, 80\% and
  95\% intervals while the orange curve shows the median forecast. The black
  curve shows ``finalized'' data, as reported in May 2021.}   
\label{fig:trajectory}
\end{figure}

@@

In addition to this pure autoregressive model, we also consider five
logistic models of the form:  
\begin{equation}
\label{eq:hotspot_ar_x}
\logit \big( \P(Z_{\ell,t+a} = 1 \ | \ Y_{\ell,s},\ X_{\ell,s},\ s \leq t) \big)
=  \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{equation}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit the above models,
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x}, using logistic regression,   
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=.5\columnwidth]{fig/leading-only-1.pdf}
  \caption{Correlation of the difference in WIS with the ``leadingness'' of the
    indicator at the target date, stratified by up, down, or flat period.} 
  \label{fig:leading}
\end{figure}

@@

\subsubsection*{Acknowledgements}
We thank Matthew Biggerstaff, Logan Brooks, Johannes Bracher, Michael  
Johansson, Evan Ray, Nicholas Reich, and Roni Rosenfeld for several
enlightening conversations about forecasting, scoring, and evaluation.  
This material is based on work supported by gifts from Facebook, Google.org,
the McCune Foundation, and Optum; Centers for Disease Control and Prevention
(CDC) grant U01IP001121; the Canadian Statistical Sciences Institute; National 
Sciences and Engineering Research Council of Canada (NSERC) grant
RGPIN-2021-02618; and National Science Foundation Graduate Research
Fellowship Program (NSF GRFP) award DGE1745016.

%\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliographystyle{unsrtnat}
\bibliography{../../common/covidcast.bib}

\clearpage

\appendix

\section*{Supplemental information}

\input{supplement-medrxiv.tex}
